SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2020-01-16 2017-Xudong-SMT/ActaInformatica.tex
Upload a document reviewer2.docx. And Typos.

\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
%\newcommand{\TODO}[1]{}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\;\land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\definecolor{darkgreen}{rgb}{0.2, 0.55, 0.2}
%\newcommand{\Simon}[1]{\textcolor{darkgreen}{#1}}
%\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
%\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
%\newcommand{\YX}[1]{{\color{red}#1}}
\newcommand{\Simon}[1]{{#1}}
\newcommand{\ERIC}[1]{{#1}}
\newcommand{\QIN}[1]{{#1}}
\newcommand{\YX}[1]{{#1}}

@@

\institute{Xudong Qin(\email{marsxd@gmail.com}) \and
  Zechen Hou (\email{zechen@sei.ecnu.edu.cn}) \and
  Yuxin Deng (\email{yxdeng@sei.ecnu.edu.cn}) \and
  Min Zhang (\email{mzhang@sei.ecnu.edu.cn}) \at
  Shanghai Key Laboratory of Trustworthy Computing, East China Normal University, China%% \\
  %% \email{marsxd@gmail.com, zechen@sei.ecnu.edu.cn, yxdeng@sei.ecnu.edu.cn, mzhang@sei.ecnu.edu.cn}
  \and
  Simon Bliudze \at
  INRIA Lille -- Nord Europe,
  40 avenue Halley,
  59650 Villeneuve d'Ascq,
  France\\
  \email{simon.bliudze@inria.fr}
  \qquad
  \orcid{0000-0002-7900-5271}
  \and
  Eric Madelaine \at
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia
  Antipolis, France,\\
  \email{eric.madelaine@inria.fr}
  \qquad
  \orcid{0000-0002-5552-5993}
}

@@

\paragraph{Contribution.}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking the satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show the correctness of the optimised algorithm w.r.t. the original one.
	% the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\ERIC{In Fig. \ref{schema:enable-pnets} we give two possible pNet  encodings of the ``Enable'' operator, \YX{which are  state-oriented and data-oriented, respectively.} 
The first one is called $EnableState$, and uses a controller $CState$ with three simple control actions, $l, r$ and $\delta$,  to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the LOTOS statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new action constant constructors $l, r, \delta$ and $acc$.}

\ERIC{This shows an example of the generic $\texttt{FUN}$ constructor used to construct expressions (of the Action sort, or of any data sort) in any algebra presentations.
  Here \texttt{FUN} will be instanciated e.g. as, $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$.
  %Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$.
  Note that the same action constructor $\delta$ is also used as a constant action in the controller $CState$.
}

\ERIC{The $EnableData$ pNet uses the same algebra presentation.}\QIN{The main difference between $EnableState$ and $EnableData$ lies in the construction of their controllers $CState$ and $CData$ that we will discuss in the next section.}
%\QIN{
%Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. As there is no more constructor required, we can use the same algebra presentation as $EnableState$ in this case.
%}
\end{example}

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and then checked for
satisfiability (function \textit{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

\QIN{As the removal of the unsatisfiable transitions \YX{may entail some transitions without preceeding} transitions, }
the set $L1$ might contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

 Let $pn$ be a pNet. The set of states 
 managed by the two algorithms both are subsets of the cartesian
 product of the PLTS states, as defined in page \pageref{def-states}. In a
 similar way, the set of state variables attached to each state are
 also equal. Comparing  
 transitions requires a little more care, because within the
 $MakeTransitions$ function, fresh variables are created everytime a
 synchronisation vector is used (this occurs within the $MkPred$ and
 $matching$ functions). However, given a global state, the call of
 $MakeTransitions$ will always return exactly the same set of OTs, up to a
renaming of these fresh variables. When this is not ambiguous, we will
speak of equality (of transitions, and of open automata) without the phrase ``up to a renaming of fresh variables''.

@@

Figure~\ref{schema:smt-lib} shows the translation of  Example~\ref{schema:unsat-ot}, page \pageref{schema:unsat-ot},
in the SMTlib syntax. It contains the declaration
of the relevant part of the LOTOS action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

@@

\paragraph{Translation of open transitions.}
In~\cite{Avocs-RR} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
	\item to collect all variables in the transition, and declare them (using \texttt{declare-const}),
	\item to check the well-formedness and correct typing of expressions,
	\item to translate the predicate into a conjunction of assertions,
	\item to translate the state-variable
	assignments into a disjunctive assertion, if the assignments are present in the source state.
\end{itemize}
%
This translation ensures that no runtime error will occur
in the SMT engine.

@@

For this example, the tool (using Algorithm~\ref{alg1}) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.
%% \noteSB{Eric, in the state 20, $b0$ should always be $false$: $\PortFail(true)$ is not possible in Control, and the safety property (\ref{eq:FM:safety}) does not hold otherwise.}

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
open automaton, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
We have used a running example based on two possible
encodings of the Enable operator of LOTOS. The automata computed with
the algorithm can be used to prove equational properties of the
operator.

Then we define a ``smart'' version of the algorithm that explores, as
much as possible, only the reachable part of the open automaton. The
handling of assignments makes the reachable computation more
challenging. We show that the result is equivalent to the original
algorithm.

>>>>

\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
%\newcommand{\TODO}[1]{}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\;\land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\definecolor{darkgreen}{rgb}{0.2, 0.55, 0.2}
%\newcommand{\Simon}[1]{\textcolor{darkgreen}{#1}}
%\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
%\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
%\newcommand{\YX}[1]{{\color{red}#1}}
\newcommand{\Simon}[1]{{#1}}
\newcommand{\ERIC}[1]{{#1}}
\newcommand{\QIN}[1]{{#1}}
\newcommand{\YX}[1]{{#1}}
\newcommand{\QINX}[1]{\textcolor{orange}{#1}}

@@

\institute{Xudong Qin(\email{52174501019@stu.ecnu.edu.cn}) \and
  Zechen Hou (\email{51174500089@stu.ecnu.edu.cn}) \and
  Yuxin Deng (\email{yxdeng@sei.ecnu.edu.cn}) \and
  Min Zhang (\email{mzhang@sei.ecnu.edu.cn}) \at
  Shanghai Key Laboratory of Trustworthy Computing, East China Normal University, China%% \\
  %% \email{marsxd@gmail.com, zechen@sei.ecnu.edu.cn, yxdeng@sei.ecnu.edu.cn, mzhang@sei.ecnu.edu.cn}
  \and
  Simon Bliudze \at
  INRIA Lille -- Nord Europe,
  40 avenue Halley,
  59650 Villeneuve d'Ascq,
  France\\
  \email{simon.bliudze@inria.fr}
  \qquad
  \orcid{0000-0002-7900-5271}
  \and
  Eric Madelaine \at
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia
  Antipolis, France,\\
  \email{eric.madelaine@inria.fr}
  \qquad
  \orcid{0000-0002-5552-5993}
}

@@

\paragraph{Contribution.}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking the satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions \QINX{during the execution of} the residual algorithm, and show the correctness of the optimised algorithm. \QINX{Meanwhile its result is equivalent to the original one.}
	% w.r.t. the original one.
	% the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\ERIC{In Fig. \ref{schema:enable-pnets} \QINX{we give the pNet encoding of the ``Enable'' operator in two different styles,} \YX{which are  state-oriented and data-oriented, respectively.} 
The first one is called $EnableState$, and uses a controller $CState$ with three simple control actions, $l, r$ and $\delta$,  to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the LOTOS statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new action constant constructors $l, r, \delta$ and $acc$.}

\QINX{It shows an example that the generic $\texttt{FUN}$ constructor can be used to construct expressions (of the Action sort, or of any data sort) in any algebra presentations.
  Here \texttt{FUN} will be \QINX{instantiated} e.g. as $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$.
  %Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$.
  Note that the same action constructor $\delta$ is also used as a constant action in the controller $CState$.
}

\ERIC{The $EnableData$ pNet uses the same algebra presentation.} \QIN{The main difference between $EnableState$ and $EnableData$ lies in the construction of their controllers $CState$ and $CData$ that we will discuss in the next section.}
\end{example}

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is \QINX{clearly} not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and then checked for
satisfiability (function \textit{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

\QIN{As the removal of the unsatisfiable transitions \YX{may entail some transitions without preceeding} transitions, }
the set $L1$ might contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$.
%as defined in page~\pageref{def-states}
More precisely, this is a residual algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

 Let $pn$ be a pNet. The set of states 
 managed by the two algorithms both are subsets of the cartesian
 product of the PLTS states.
 %, as defined in page \pageref{def-states}. 
 In a similar way, the set of state variables attached to each state are
 also equal. Comparing  
 transitions requires a little more care, because within the
 $MakeTransitions$ function, fresh variables are created everytime a
 synchronisation vector is used (this occurs within the $MkPred$ and
 $matching$ functions). However, given a global state, the call of
 $MakeTransitions$ will always return exactly the same set of OTs, up to a
renaming of these fresh variables. When this is not ambiguous, we will
speak of equality (of transitions, and of open automata) without the phrase ``up to a renaming of fresh variables''.

@@

Figure~\ref{schema:smt-lib} shows the translation of  Example~\ref{schema:unsat-ot} %,page \pageref{schema:unsat-ot},
in the SMTlib syntax. It contains the declaration
of the relevant part of the LOTOS action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

@@

\paragraph{Translation of open transitions.}
In~\cite{Avocs-RR} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
	\item to collect all variables in the transition, and declare them (using \texttt{declare-const}),
	\item to check the well-formedness and correct typing of expressions,
	\item to translate the predicate into a conjunction of assertions,
	\item to translate the state-variable
	assignments into a disjunctive assertion, if the assignments are present in the source state.
\end{itemize}
%
This translation ensures that \QINX{the error occurred will not be caused by the incorrect inputs.}
%no runtime error will occur in the SMT engine.

@@

For this example, the tool (using Algorithm~\ref{alg1}) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3. \QINX{According to the unsatisfiable transitions, there are 3 states which are not reachable.} The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.
%% \noteSB{Eric, in the state 20, $b0$ should always be $false$: $\PortFail(true)$ is not possible in Control, and the safety property (\ref{eq:FM:safety}) does not hold otherwise.}

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
open automaton, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates\QINX{, and then we submit them to the Z3 solver.}
%
%\hl{%
%
We have used a running example based on two possible
encodings of the Enable operator of LOTOS. The automata computed with
the algorithm can be used to prove equational properties of the
operator.

Then we define a ``smart'' version of the algorithm that explores, as
much as possible, only the reachable part of the open automaton. The
handling of assignments makes \QINX{the computation of the reachable states} more
challenging. We show that the result is equivalent to the original
algorithm.
